---
title: "Exposure-response modeling for binary data"
output:
  slidy_presentation:
    theme: cerulean
    fig_width: 9
    fig_height: 5
    font_adjustment: 3
    transition: faster
    css: slidystyles.css
    footer: tunka &copy 2023
    mathjax: local
    self_contained: false
    df_print: tibble
#bibliography: alpha.bib
#nocite: | 
editor_options: 
  chunk_output_type: console
---


# Outline

```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
library(tidyverse)
library(stringr)
library(haven)
library(survival)
library(GGally)
library(mgcv)
library(binom)
library(texreg)
library(DHARMa)
library(brms)
library(ggplot2)
#library(pmtables)

theme_set(theme_bw())
``` 

* Encountering binary data
* Groundwork
   * Notational conventions
   * Modeling from a probabilistic viewpoint
   
* Bernoulli distribution
   * Defining models
   * Maximum likelihood

* Visualizing relationships
* Models, model checking, and measures of effect
  

# Binary Data Example

* This data set is comprised of a two-week study (protocol A) and a six-week study (protocol B)

* The data set includes patients of type `PT2`, a patient type that (let's say) has not been studied at higher dose levels.

* Suppose further that a dose level under consideration for a phase 2 study in `PT2` would have typical value exposures near 2.5 ug/mL.

* The objective of *this* analysis will be to determine whether the Adverse Event (AE) rate will be "sufficiently low" in `PT2` at that exposure.

# Data snapshot A

Pharmacology data:

* 210 observations with 11 variables
* 180 people

<!--- PopPK is the study of variability in drug concentrations within a patient population receiving clinically relevant doses of a drug of interest. --->


```{r, echo=FALSE}
#source of data:
setwd("C:/Users/ruben/Documents/2023/Academics/Time-to-Event-Analysis_Course-01")
```

```{r}
load('data/aedat.RDS')
```

Variables and possible values:

- STUDYID: Study [PROTA and PROTB]
- STUDYDUR: Duration of Observation (weeks) [2 and 6]
  - PROTA = 2 weeks (69 cases)
  - PROTB = 6 weeks (141 cases)
- USUBJID: Subject ID [UID-001, UID-002,...]
- SEXTXT: Sex [FEMALE and MALE]
- PTTYPE: Patient Type [HV, PT1 and PT2] -the reference level is "HV"
- PBO: Placebo/Treatment [PBO and TRT]
- CAVGSS: Time-averaged steady-state concentration (ug/mL) [0, 1.98, ...]
- BWT: Baseline Body Weight (kg) [77.4, 83, ...]

- AETOXGR: Adverse Event Grade [0,1,2 and 3]
- AE01: Whether Severe Adverse Event Occured [0 and 1]
  - AE01 = 1 <=> AETOXG = 3
- TTE: Time of Adverse Event or Censoring Event (weeks) [0.0008,...]


```{r}
str(aedat)
```

# Data snapshot B

First 10 rows (without showing SEXTXT):

```{r}
aedat[,-2] %>% head(10)
```

# Data snapshot C

Some tables and plots:

```{r}
table(aedat[,c('STUDYID','STUDYDUR')])
table(aedat[,c('STUDYID','PTTYPE')])
table(aedat[,c('STUDYID','PBO')])
table(aedat[,c('PBO','PTTYPE')])
```

```{r}
table(aedat[,c('AETOXGR','AE01')])
table(aedat[,c('STUDYID','AE01')])
table(aedat[,c('PTTYPE','AE01')])
```

```{r}
data <- aedat
data$AE01 <- factor(data$AE01)
ggplot(data, aes(x = AE01, y = CAVGSS)) +
     geom_boxplot() +
     labs(title = "Boxplot by AE01",
          x = "AE01",
          y = "CAVGSS")

data <- aedat
ggplot(data, aes(x = PTTYPE, y = CAVGSS)) +
     geom_boxplot() +
     labs(title = "Boxplot by PTTYPE",
          x = "PTTYPE",
          y = "CAVGSS")

data <- aedat
ggplot(data, aes(x = PBO, y = CAVGSS)) +
     geom_boxplot() +
     labs(title = "Boxplot by PBO",
          x = "PBO",
          y = "CAVGSS")

data <- aedat
ggplot(data, aes(x = STUDYID, y = CAVGSS)) +
     geom_boxplot() +
     labs(title = "Boxplot by PBO",
          x = "STUDYID",
          y = "CAVGSS")

```

# Transforming the data

* Let's factorize the variable AETOXGR: None, Mild, Moderate and Severe
 
* Let's create the variable TTE_SEVERE to show the time when AETOXGR = Severe

```{r, echo=FALSE}
aedat <- aedat %>% 
  mutate(AETOXGR = factor(AETOXGR, 0:3, labels=c("None","Mild","Moderate","Severe"))) %>% 
  group_by(USUBJID) %>%
  # End of study for patients without a severe event
  mutate(TTE_SEVERE = case_when(
    STUDYID=="PROTA" ~ 2,
    STUDYID=="PROTB" ~ 6),
    # Time of severe event for those that had one
    TTE_SEVERE = ifelse(AETOXGR=="Severe", TTE, TTE_SEVERE)
  ) %>% 
  ungroup()
aedat[,c(-2,-4)] %>% head(10)
```

* Let's reduce the data to have only one observation per ID. We will use the min value of TTE_SEVERE variable for every ID that has more than one observation

```{r, echo=FALSE}
aedat <- aedat %>% 
  group_by(USUBJID) %>% 
  arrange(USUBJID, TTE_SEVERE) %>% slice(1) %>%
  ungroup()
aedat[,c(-2,-3,-4)]  %>% head(10)
dim(aedat)
```

Pharmacology data:

* 180 observations or 180 people

* Now let's create the variable Quartile to indicate the quartile for the variable CAVGSS only when PBO is  Treatment [TRT]

```{r, echo=FALSE}
aedat <- aedat %>% 
  group_by(PBO) %>%
  mutate(Quartile = ifelse(PBO == "PBO", "PBO",
                           paste0("Q", ntile(CAVGSS, n = 4)))) %>% 
  ungroup()
aedat[,c(-2,-3,-4)]  %>% head(10)
```

As an option to analize, we could normalize some variables. For example the BWT:

* Let's create a normalized variable BWT_norm of the Body weight BWT (using the median)

```{r}
# Median(BWT):
ref_bwt <- median(aedat$BWT)
ref_bwt
```

Using the median in order to calculate BWT_norm
```{r, echo=FALSE}
dat_mod <- aedat %>% mutate(
  BWT_norm = BWT - ref_bwt,
  PTTYPE = factor(PTTYPE, levels = c("PT1", "PT2", "HV"))
)
dat_mod[,c(-2,-3,-4, -8)] %>% head(10)
```

# Data of interest:

Let's return to the previous data with original BWT:

```{r, echo=FALSE}
aedat %>% 
  mutate(across(where(is.numeric), pmtables::sig)) %>% 
  select(STUDYID, USUBJID, PBO, CAVGSS, AE01, AETOXGR) %>% 
  head(n=10)
```
We can find the reference for pmtables here: https://metrumresearchgroup.github.io/pmt-book/stable.html

* `STUDYID`: Study Protocol [PROTA and PROTB]
* `USUBJID`: Unique subject id
* `AE01`: [1 or 0] Adverse event of grade 3 or higher

# Visualizing relationships

Objective:

> Plot the probability of a grade 3 or higher AE vs predictor 

Types of plots will depend on the predictor variable:

  * Categorical variable
  * Continuous variable

# Relationship with categorical variable

* x-axis: Categorical variable [`STUDYID`]
* y-axis: Probability of AE [`AE01`]
* geom: `point`

# Probability vs categorical variable - Simple method: a point

Using the predictor `STUDYID`:
```{r}
freq_table <- table(aedat[,c('STUDYID','AE01')])
freq_table
row_sums <- rowSums(freq_table)
freq_table / row_sums
```
We can conclude that the last column give us the probability for event grade 3 or higher

# Probability vs categorical variable {.sourceCode.r.small}
::: columns

:::: column
```{r, 'cat-plot', fig.show='hide'}
aedat %>% 
  group_by(STUDYID) %>% 
  summarise(x = sum(AE01), n=n()) %>% 
  mutate(phat = x / n,
         lcl = binom.confint(x,n,methods = "wilson")$lower,
         ucl = binom.confint(x,n,methods = "wilson")$upper) %>% 
  ggplot(aes(x=STUDYID, y=phat)) + 
  geom_point() +
  geom_errorbar(aes(ymin=lcl, ymax=ucl), 
                width = 0.2) +
  ylim(0,1) +
  labs(x='Study', y='Probability of severe AE')
```

Note: `binom.confint` is in the `binom` package

::::

:::: column
```{r, ref.label='cat-plot', echo=FALSE, out.width="95%"}

```
::::

:::


# Probability vs a continuous variable - Simple plots: 
x = CAVGSS

```{r}
plot(aedat$CAVGSS,aedat$AE01, xlab = "CAVGSS", ylab = "Event AE01 = 0 or 1")

data <- list(
  AE01_0 = aedat$CAVGSS[aedat$AE01 == 0],
  AE01_1 = aedat$CAVGSS[aedat$AE01 == 1])

# Create a horizontal boxplot
boxplot(data, horizontal = TRUE, main = "Horizontal Boxplot Example",
        xlab = "CAVGSS", ylab = "Event AE01 = 0 or 1")
```


# Probability vs a continuous variable {.sourceCode.r.small}

We are going to show the curves to estimate the proportions by CAVGSS and we also divide the data into Quartiles in order to give the proportion of AE01 = 1 in every case (Quartile).  

:::{.columns}

::::{.column}
```{r}
dat_eda <- aedat %>% group_by(Quartile) %>% mutate(MedConc = median(CAVGSS))
```

```{r, 'cont-er', fig.show='hide'}
ggplot() +
  # Add tick marks at top and bottom
  geom_rug(data = filter(dat_eda, AE01 == 0),
           aes(x = CAVGSS), sides = "b") +
  geom_rug(data = filter(dat_eda, AE01 == 1),
           aes(x = CAVGSS), sides = "t") +
  # Add smooth mean function.   Note: Use of gam with 
  #  binomial family to keep predictions on (0,1) scale
  geom_smooth(data = dat_eda,
              aes(x = CAVGSS, y = AE01),
              method='gam', formula=y~s(x),
              method.args = list(family='binomial')) +
  # Add points and CIs
  stat_summary(
    data = dat_eda,
    aes(x = MedConc, y = AE01), #, group = MedConc),
    fun = function(y) sum(y) / length(y),
    geom = "point"
  ) +
  stat_summary(
    data = dat_eda,
    aes(x = MedConc, y = AE01),   
    # Wilson CI recommended by Agresti and Coull (2000) review paper
    fun.min = function(y) {
      binom.confint(sum(y),length(y),
                    methods = "wilson")$lower
      },
    fun.max = function(y) {
      binom.confint(sum(y), length(y),
                    methods = "wilson")$upper
      },
    geom = "errorbar"
  ) +
  labs(x='Steady-state Cavg', y='Probability of severe AE')
```
::::

::::{.column}
```{r, ref.label='cont-er', echo=FALSE}

```

::::

:::


# Plotting tips

* Break exposure into quartiles, compare incidence rate across quartiles
  * Textbook variance formula $\frac{p(1-p)}{n}$ isn't the best choice with small sample sizes and probabilities near zero or one
  * Recommendation: Use Wilson interval (implemented in  `binom.confint`)
* Rug plots (`geom_rug`)
* Stratify by other covariates of interest


# Workbook 02: Visualizing binary data
